---
title: "Van Analysis"
author: "Kelvin N. Kinyanjui"
date: '2022-08-17'
output:
  html_document:
    df_print: paged
  pdf_document: default
   
---

# Business Problem   
***

The client owns a business that maintains a lot of vans.
==he client provided an spreadsheet file with *Engine Type*,*Mileage*, *Service Date*, *Cost of repair*, and *Van Number*.[Shared File](https://docs.google.com/spreadsheets/d/11zPiOVSR7AlZEzn0jGHPik0ISszIb5w7rFvrMT7P8pQ/edit?usp=sharing)  

**Requirements**:  
1. Visualize data to highlight important insights on engine service per engine and where possible per van  
2. Visualize data to determine at what point the vans are too expensive to maintain.  

## Executive Summary

- Generally, the number of service times increase as the mileage increases.   
- Therefore, an increase in service time reduces the efficiency of the vehicles  
- Most service engine costs are below $1000, with the highest costs being recorded for vans with Promaster 3.6L engine  
- Generally, vans engines are efficient around 100,000 miles.   
- Past the 100,000 miles mark, most vans become expensive as service costs exponentially increases.   
- Vans with GC 3.6L engine are the most efficient and cheap to maintain.   
- On the other hand, vans with promaster 3.6L engine are the least efficient and the most expensive to maintain. 

### Conclusion and Data Limitations  

Most vans’ engines past 100,000 miles are expensive to maintain and should be replaced with new engines  
**Client Implications**:  
- This conclusion should be acted upon with caution due to the data limitations.   

**Data Limitations**  
- *Poor quality data*: The data had gross inconsistencies in mileage when cross checked against service dates  
- *Inadequate data*:The number of vans per engine varied greatly with varying mileage ranges which would make generalizations difficult.  


# Data Preparation, Manipulation, and Analysis   
*** 

## Setting Up my environement  



Loading vans data from Google sheets using googlesheets4::read_sheet() function

```{r loading_data}
raw_data <- readxl::read_xlsx("Van_Mileage.xlsx")
head(raw_data)
```

Renaming column heads with readable words

```{r change column names}
#assign the raw data to a new variable
van_data <- raw_data 

#use a janitor package function to clean column names ensuring column names contains only numbers, characters & letters
van_data <- janitor::clean_names(van_data)

#view the column names
colnames(van_data)

#change the column names to be more descriptive
van_data <- dplyr::rename(van_data, engine_type = "veh_type", service_date = "date", total_cost = "total")
head(van_data)

```

Summarizing data to get an overview and while evaluating for inconsistencies

```{r summarize_data}
#sorting van_data in ascending order using mileage
van_data <- dplyr::arrange(van_data, mileage)
head(van_data)

# The nested code below can also be expressed using pipes

# vans_per_engine <- van_data %>% 
#   dplyr::group_by(engine_type) %>% 
#   dplyr::summarize(max_mileage = max(mileage), min_mileage = min(mileage), unique_vans = dplyr::n_distinct(van_number))

#getting a high level summary of the data
vans_per_engine_type <- dplyr::summarize((dplyr::group_by(van_data, engine_type)), max_mileage = max(mileage), min_mileage = min(mileage), unique_vans = dplyr::n_distinct(van_number))
head(vans_per_engine_type)

#Getting a summary for each van
# vans_summary <- van_data %>% 
#   dplyr::group_by(engine_type, van_number) %>% 
#   dplyr::summarize(max_mileage = max(mileage), min_mileage = min(mileage), number_of_service_times = dplyr::n_distinct(service_date), total_service_cost = sum(total_cost) )
# View(vans_summary)
# head(vans_summary)

vans_summary <- dplyr::summarize((dplyr::group_by(van_data, engine_type, van_number)), max_mileage = max(mileage), min_mileage = min(mileage), number_of_service_times = dplyr::n_distinct(service_date), total_service_cost = sum(total_cost) )

head(vans_summary)
View(vans_summary)
```

## Data Cleaning and Data Validation   

From the summary above, there are noticeable inconsistencies in the data.  
1. Inconsistent naming in the van_number column  
2. There are 32 observations despite the last van being listed as van 30. After evaluation there are two Van9s (Van N09 and Van 09) and two Van14s (Van 14 &Van N14).   

In this case, visualizing the two observations: van9s and van14s.  

Filtering data to only show the two variations of van 09 and Van 14.  

```{r compare van 9 & 14, warning=FALSE}
 # Filtering the data
van_9s <- dplyr::filter(van_data, van_number== "VAN 09" | van_number== "VAN N09")
van_14s <- dplyr::filter(van_data, van_number== "VAN 14" | van_number== "VAN N14")

head(van_9s)
View(van_9s)#From this view one can tell that the vans are different because they have different engine types 

head(van_14s)
View(van_14s)
#dataframes for the ggplot2 plots

#1. date_breaks for setting the breaks in the x axis dates
date_breaks <- seq(as.POSIXct("2019-06-01"), as.POSIXct("2022-06-01"), by = "6 month")

#plotting a scatter plot to visualize the data(van_9s)
ggplot2::ggplot(van_9s) + 
  ggplot2::geom_point(mapping = ggplot2::aes(x= service_date, y= mileage, color = van_number, shape= van_number)) +
  ggplot2::theme_bw()+
  ggplot2::labs(title = "Mileage vs Service Date", subtitle = "Visualizing VAN 09 and VAN N09 to check for similarities or differences", caption = "The two vans are different" ) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle =25, hjust = 1))

#plotting a scatter plot to visualize the data(van_14s).

ggplot2::ggplot(van_14s) + 
  ggplot2::geom_point(mapping = ggplot2::aes(x= service_date, y= mileage, color = van_number, shape= van_number)) +
  ggplot2::theme_bw() + 
  ggplot2::labs(title = "Mileage vs Service Date", subtitle = "Visualizing VAN 14 and VAN N14 to check for similarities or differences", caption = "The two vans are different" ) + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle =25, hjust = 1)) + 
  ggplot2::scale_x_datetime(breaks = date_breaks)+ 
  ggrepel::geom_text_repel(ggplot2::aes(service_date, mileage, label= mileage), size=2)+
ggplot2::annotate("rect", xmin = as.POSIXct("2020-03-01"), xmax= as.POSIXct("2020-11-01"), ymin = 175000, ymax = 220000, alpha = .1, fill = "blue" ) +
  ggplot2::annotate("text", x= as.POSIXct("2020-11-15"), y= 160000, label = "The last two scatter points are \n inconsistent with the other scatter points", size =3)+
  ggplot2::annotate("text", x= as.POSIXct("2020-11-15"), y= 25000, label = "Some values of VAN N14 might \n have been recorded as VAN 14", size =3)

```

From the plot above, a number of things are ensued.  
1. There are inconsistencies in the service_dates and the recorded mileage (Will be handled in the next step)  
2. VAN N09 can be changed to van 31  
3. VAN N14 can be changed to VAN 14 to VAN 32  
4. Client gave a go ahead on changing the VAN 14 observations with a mileage below 50000 to Van32  

### Changing Observations(van_number) for consistency     

First we need to replace VAN N09 with VAN 31 and VAN 14 with VAN32  

```{r strings cleaning}
#assigning van_data to another variable.  
# The code below modifies the original data frame. Therefore, it is wise to assign it to another variable 
van_data_32 <- van_data

#modifying the observations that were recorded incorrectly 
van_data_32$van_number[van_data_32$van_number=="VAN 14"& van_data_32$mileage < 50000] <- "VAN N14"

#modifying van numbers for two vans
van_data_32$van_number[van_data_32$van_number=="VAN N09"] <- "VAN 31"
van_data_32$van_number[van_data_32$van_number=="VAN N14"] <- "VAN 32"

#visualizing the two  changed van numbers
# testvan <-dplyr::filter(van_data_32, van_number== "VAN 09" | van_number== "VAN 31")
# testvan2 <-dplyr::filter(van_data_32, van_number== "VAN 14" | van_number== "VAN 32")
# ggplot2::ggplot(data =testvan ) + ggplot2::geom_point(mapping = ggplot2::aes(x= service_date, y=mileage,color = van_number))
# ggplot2::ggplot(data =testvan2 ) + ggplot2::geom_line(mapping = ggplot2::aes(x= service_date, y=mileage,color = van_number)) +ggplot2::facet_wrap(~engine_type)

#changing van_numbers names for consistency. 
van_data_32$van_no <- stringr::str_sub(van_data_32$van_number, -2)
van_data_32$van <- stringr::str_c("van_", van_data_32$van_no)
van_data_32 <- dplyr::select(van_data_32, van, engine_type, mileage, service_date, total_cost)
head(van_data_32)
#View(van_data_32)

```


### Data Validation    

It is expected that an increase in mileage should be accompanied by a progression in Date.  

Therefore, we will make plots per van for mileage against service date to verify if the data is correct and detect any outliers.   

```{r data validation}

#Creating a broad visualization for all the vans

#ggplot2::ggplot(data =van_data_32 ) + ggplot2::geom_line(mapping = ggplot2::aes(x= service_date, y=mileage,color = van)) +ggplot2::facet_wrap(~engine_type)

# We can filter the data according to engine (combine engine_types with few vans)

# gc_3.6l <- dplyr::filter(van_data_32, engine_type== "GC 3.6L")
# promaster <- dplyr::filter(van_data_32, engine_type== "PROMASTER 3.6L")
# other_vans <- dplyr::filter(van_data_32, engine_type != "GC 3.6L" & engine_type != "PROMASTER 3.6L")

# ggplot2::ggplot(data =gc_3.6l, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point()  
# 
# ggplot2::ggplot(data =promaster, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() 
# 

# ggplot2::ggplot(data =other_vans, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() 

#deleting a row that does not meet a specific condition
#van_data_32 <- van_data_32[!(van_data_32$van == "van_02" & van_data_32$mileage == 16699),]

#van_data_32$service_date[van_data_32$mileage == 19000 & van_data_32$van == "van_06"] <- as.POSIXct("2021-12-06")


van_data_32 <- dplyr::arrange(van_data_32, van, mileage)
#View(van_data_32)
```


#### Removing Outliers(out of range) van_02  
Some of the observations that were changed:    

| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|
|van_02|2020_11_23   |16699      |removed      |removed    |-129.04             |
|van_02|2020_12_23   |213404     |not_changed  |213440     |572.87              |

```{r cleaning van 2}

#filtering data for each van, checking for outliers through a visualization, confirming with the client and modifying the values accordingly

#van_data_32 <- dplyr::arrange(van_data_32, van, mileage)
#View(van_data_32)

#make changes and then filter the data out to verify that the changes have been effected



head(van_data_32)

 select_van <- dplyr::filter(van_data_32, van== "van_02")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point()+ggplot2::labs(title = "Mileage vs Service Date: van_02 with outlier(s)")

 #deleting a row that does not meet a specific condition
 
van_data_32 <- van_data_32[!(van_data_32$van == "van_02" & van_data_32$mileage == 16699),]

select_van_clean <- dplyr::filter(van_data_32, van== "van_02")
ggplot2::ggplot(data = select_van_clean, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +ggplot2::labs(title = "Mileage vs Service Date: van_02 without outlier(s)")

```


#### Removing Outliers(out of range) in van_12     

Some of the observations that were changed include:    
| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|   
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|  
|van_12|2021_02_28   |292289     |removed      |removed    |-58.14             |  


```{r cleaning van_12}
 
 select_van <- dplyr::filter(van_data_32, van== "van_12")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +ggplot2::labs(title = "Mileage vs Service Date: van_12 with outlier(s)")


 #deleting a row that does not meet a specific condition
 
van_data_32 <- van_data_32[!(van_data_32$van == "van_12" & van_data_32$mileage == 292289),]

select_van_clean <- dplyr::filter(van_data_32, van== "van_12")
ggplot2::ggplot(data = select_van_clean, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() + ggplot2::labs(title = "Mileage vs Service Date: van_12 without outlier(s)")
```

   
#### Removing Outliers(out of range) from van_16  

Some of the main changes included:    

| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|
|van_16|2021-10-15   |511155     |removed      | removed   |-97.28              |


```{r cleaning van_16}


select_van <- dplyr::filter(van_data_32, van== "van_16")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +ggplot2::labs(title = "Mileage vs Service Date: van_16 with outlier(s)")


#  #deleting a row that does not meet a specific condition
 van_data_32 <- van_data_32[!(van_data_32$van == "van_16" & van_data_32$mileage == 511155),]
 
 select_van_clean <- dplyr::filter(van_data_32, van== "van_16")
 ggplot2::ggplot(data = select_van_clean, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() + ggplot2::labs(title = "Mileage vs Service Date: van_16 without outlier(s)")

```



#### Cleaning van_17   

Some of the main changes included:    

| Van  |Date         |mileage    |new_date     |new_mileage| total_cost|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:---------:|
|-     |-            |-          |-            |-          |-          |

van_17 is a van of interest. Despite having an inconsistency, the values are within the range. Therefore, the inconsistency will not be removed.

```{r cleaning van_17 }

select_van <- dplyr::filter(van_data_32, van== "van_17")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +  ggplot2::labs(title = "Mileage vs Service Date: van_17 cleaning")

```
  
  
#### Removing Outliers (out 0f range) from Van_19   

Some of the main changes included:   

| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|
|van_19|2021-11-03   |26501      |removed      |remove     |-419.54             |
             |


```{r cleaning van_19}


select_van <- dplyr::filter(van_data_32, van== "van_19")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +ggplot2::labs(title = "Mileage vs Service Date: van_19 with outlier(s)")


#  #deleting a row that does not meet a specific condition
 van_data_32 <- van_data_32[!(van_data_32$van == "van_19" & van_data_32$mileage == 26501),]
 
 select_van_clean <- dplyr::filter(van_data_32, van== "van_19")
 ggplot2::ggplot(data = select_van_clean, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() + ggplot2::labs(title = "Mileage vs Service Date: van_19 without outlier(s)")

```


   
#### Removing Outliers (out of range) from van 26   

Some of the main changes included:     
| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|
|van_26|2021-08-02   |184900     |removed      |removed    | 67.40             |




```{r Cleaning van_26}

select_van <- dplyr::filter(van_data_32, van== "van_26")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() +ggplot2::labs(title = "Mileage vs Service Date: van_26 with outlier(s)")


#  #deleting a row that does not meet a specific condition
 van_data_32 <- van_data_32[!(van_data_32$van == "van_26" & van_data_32$mileage == 184900),]
 
 select_van_clean <- dplyr::filter(van_data_32, van== "van_26")
 ggplot2::ggplot(data = select_van_clean, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() + ggplot2::labs(title = "Mileage vs Service Date: van_19 without outlier(s)")
 
```

     
#### Cleaning van_27 (van of intrest)   

Some of the main changes included:   

| Van  |Date         |mileage    |new_date     |new_mileage| service_cost_change|
|:----:|:-----------:|:---------:|:-----------:|:---------:|:------------------:|
|


```{r cleaning van-27}


select_van <- dplyr::filter(van_data_32, van== "van_27")
 ggplot2::ggplot(data = select_van, ggplot2::aes(x= service_date, y=mileage,color = van) ) + ggplot2::geom_line() + ggplot2::geom_point() + ggplot2::labs(title = "Mileage vs Service Date: van_27")


```
   
    
## Van Data Summarization

This section will deal with summarizing the van-32 data.  

The data will be summarized per van and per engine type. 

```{r data summarization}

#Getting a summary for each van
# summary_per_van <- van_data_32 %>%
#   dplyr::group_by(van, engine_type) %>%
#   dplyr::summarize(min_mileage = min(mileage), max_mileage = max(mileage),  number_of_service_times = dplyr::n_distinct(service_date), min_service_cost = min(total_cost), max_service_cost= max(total_cost), average_cost_per_service = mean(total_cost), total_service_cost = sum(total_cost) )


summary_per_van <- dplyr::summarize((dplyr::group_by(van_data_32, van, engine_type)), min_mileage = min(mileage), max_mileage = max(mileage),  number_of_service_times = dplyr::n(), min_service_cost = min(total_cost), max_service_cost= max(total_cost), average_cost_per_service = mean(total_cost), total_service_cost = sum(total_cost) )

#View(summary_per_van)
head(summary_per_van)


# Getting a summary for each engine 
# summary_per_engine <- van_data_32 %>%
#   dplyr::group_by(engine_type) %>%
#   dplyr::summarize(number_of_vans= dplyr::n_distinct(van), min_mileage = min(mileage), max_mileage = max(mileage),  number_of_service_times = dplyr::n(), service_times_per_van = ceiling(dplyr::n()/dplyr::n_distinct(van)), min_service_cost = min(total_cost), max_service_cost= max(total_cost), average_cost_per_service = mean(total_cost), average_service_cost_per_van = round((sum(total_cost)/dplyr::n_distinct(van)),2), total_service_cost = sum(total_cost) )


summary_per_engine <- dplyr::summarize((dplyr::group_by(van_data_32, engine_type)), number_of_vans= dplyr::n_distinct(van), min_mileage = min(mileage), max_mileage = max(mileage),  number_of_service_times = dplyr::n(), service_times_per_van = ceiling(dplyr::n()/dplyr::n_distinct(van)), min_service_cost = min(total_cost), max_service_cost= max(total_cost), average_cost_per_service = mean(total_cost), average_service_cost_per_van = round((sum(total_cost)/dplyr::n_distinct(van)),2), total_service_cost = sum(total_cost) )

#View(summary_per_engine)
head(summary_per_engine)

```

     
## Finding Cumulative Sum for Total Cost. 

Finding the cumulative sum per van, engine, and overall.

```{r cumulative sum}

#first we sort the data according to mileage

van_data_final <- dplyr::arrange(van_data_32 ,mileage)

#adding a column for cumulative service cost. One can use the mutate functioning the dplyr package

van_data_final$overall_cum_cost = cumsum(van_data_final$total_cost)

# adding cumulative sum for each engine type

van_data_final <- dplyr::mutate(dplyr::group_by(van_data_final, engine_type), cum_cost_per_engine = cumsum(total_cost))

#adding cumulative sum for each van

van_data_final <- dplyr::mutate(dplyr::group_by(van_data_final, van), cum_cost_per_van = cumsum(total_cost))

#View(van_data_final)
head(van_data_final)

```



# Data Visualization   
***



## Visualizing Summary Data Per Engine   

In this section, we will need to visualize:  
- The number of vans per engine  
- The average service times per van per engine
- The mileage range per engine  
- The average service cost per van  

```{r visualizing data summary per engine}

# First we assign the summary_per_engine to a new variable. 

engine_summary <- summary_per_engine 
#head(engine_summary)

# Create the four plots for engine summaries 

# Number of vans per engine
#p1 <- 
ggplot2::ggplot(data = engine_summary, ggplot2::aes(x= number_of_vans, y = engine_type) )+
  ggplot2::geom_col() +
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Vans per engine") + 
  ggplot2::geom_text(ggplot2::aes(label = number_of_vans), hjust = 1.5, size = 3, color = "white") + 
  ggplot2::scale_y_discrete(limits = rev(c("GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.7L", "TRANSIT 3.5L", "E150 4.6L", "VOYAGER 3.6L")))                
ggplot2::ggsave("number_of_vans_per_engine.png", width = 1624, height = 900, units = "px")

# Number of service times per van per engine
#p2 <-
ggplot2::ggplot(data = engine_summary, ggplot2::aes(x=service_times_per_van, y = engine_type) )+
  ggplot2::geom_col() +
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Average service times per van")+ 
  ggplot2::geom_text(ggplot2::aes(label = service_times_per_van), hjust = 1.5, size = 3, color = "white") +
  ggplot2::scale_y_discrete(limits = rev(c("GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.7L", "TRANSIT 3.5L", "E150 4.6L", "VOYAGER 3.6L")))
ggplot2::ggsave("average service_times_per_engine.png", width = 1624, height = 900, units = "px")

 #Average service cost per van
#p4 <- 
ggplot2::ggplot(data = engine_summary, ggplot2::aes(x= average_service_cost_per_van, y = engine_type) )+
  ggplot2::geom_col() +
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Average service cost per van")+ 
  ggplot2::geom_text(ggplot2::aes(label = average_service_cost_per_van), hjust = 1, size = 2, color = "white")+ggplot2::scale_y_discrete(limits = rev(c("GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.7L", "TRANSIT 3.5L", "E150 4.6L", "VOYAGER 3.6L")))
ggplot2::ggsave("average_service_cost_per_engine.png", width = 1624, height = 900, units = "px")

#A dumbbell chart showing mileage ranges 

#p3 <- 
ggplot2::ggplot  (data = engine_summary, ggplot2::aes(x= max_mileage, y = engine_type))+
  ggplot2::geom_segment(ggplot2::aes(x= min_mileage, xend = max_mileage, y= engine_type, yend = engine_type)) +
  ggplot2::geom_point(size = 3) + 
  ggplot2::geom_point(ggplot2::aes(x= min_mileage, y = engine_type), size = 3) + 
  ggplot2::labs(title = "Mileage range per engine") + 
  ggplot2::theme_bw() +
  ggplot2::scale_x_continuous(breaks = seq(0, 500000, 50000)) + 
  ggplot2::xlab("Min_mileage - Max_mileage") +
  ggplot2::scale_y_discrete(limits = rev(c("GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.7L", "TRANSIT 3.5L", "E150 4.6L", "VOYAGER 3.6L")))
ggplot2::ggsave("mileage_range_per_engine.png", width = 1624, height = 900, units = "px")

# install and load patchwork
#p1 + p2+ p3+ p4
#ggplot2::ggsave("vans_service_summary_per_engine.png", width = 2606, height = 1608, units = "px")

```




## Visualizing Summary Data per van  

The summary data will be visualized using a Cleveland dot plot. 

A Cleveland dot plot functions similar to bar chart but it reduces the visual clatter

```{r visualizing data summary per van}

#Sort the van names according to engine type and then by total cost. 

sorted_van_names <- summary_per_van$van[order(summary_per_van$engine_type, summary_per_van$total_service_cost)]

# We assign the summary_per_van to another plot specifically for this plot
#Then turn the van column into a factor with levels in the order of the sorted_van_names

vans_summary_plot1 <- summary_per_van
vans_summary_plot1$van <- factor(vans_summary_plot1$van, levels = sorted_van_names)

#Add a variable for shapes(Can be easily done using the mutate function)

vans_summary_plot1$service_cost_range = dplyr::case_when(
  vans_summary_plot1$total_service_cost <= 10000 ~ "Less than 10,000",
  vans_summary_plot1$total_service_cost > 10000 & vans_summary_plot1$total_service_cost <= 20000  ~ "Between 10,000 and 20,000",
  vans_summary_plot1$total_service_cost > 20000 & vans_summary_plot1$total_service_cost <= 30000  ~ "Between 20,000 and 30,000",
  vans_summary_plot1$total_service_cost > 30000 & vans_summary_plot1$total_service_cost <= 40000  ~ "Between 30,000 and 40,000") 
ggplot2::ggplot(data = vans_summary_plot1, ggplot2::aes(x= van, y = max_mileage,size = total_service_cost, color= engine_type)) +  #specifies the aesthetics and data
  ggplot2::geom_segment(ggplot2::aes(xend = van), yend = 0, color = "grey50", size = 0.5) + 
  ggplot2::geom_point() +  #creates a normal scatter plot
  ggplot2::geom_point(mapping = ggplot2::aes(x= van, y= min_mileage))+
  #ggplot2::scale_size_area(max_size = 8) + 
  #ggplot2::scale_shape_manual(values = c(18, 16, 17, 15,7)) +
  ggplot2::theme_bw()+ 
  ggplot2::theme(
    #panel.grid.major.y = ggplot2::element_blank(), 
    #panel.grid.minor.y = ggplot2::element_blank(),
    #legend.position = "bottom",
  
    panel.grid.major.x = ggplot2::element_blank(), 
    axis.text.x = ggplot2::element_text(angle = 60, hjust = 1)
  ) + #removes the grid line on the y axis and modify the text on the x-axis
  #ggplot2::guides(color = ggplot2::guide_legend(title.position = "top", nrow = 3, ncol = 2, keyheight = 0.2)) +
  #ggplot2::guides(shape = ggplot2::guide_legend(title.position = "top", nrow = 3, ncol = 2, keyheight = 0.2)) +
  #ggplot2::guides(size = ggplot2::guide_legend(title.position = "top", nrow = 3, ncol = 2, keyheight = 0.1))+
  ggplot2::guides(size = "none")+
  ggplot2::ylab("Min_mileage - Max_mileage")+
  ggplot2::xlab("Vans ordered per engine and \n per increasing total service cost")+
  ggplot2::scale_y_continuous(breaks = seq(0, 500000, 50000))+  # make the y axis continuous
  ggplot2::labs(
    title = "Vans Service Summary", 
    subtitle = "The visualization shows the mileage range, the total service cost range, and engine type per van")
  #ggrepel::geom_text_repel(ggplot2::aes(van, max_mileage, label= number_of_service_times, angle =60 , vjust = 0.5, hjust = 0.5), size=3)

ggplot2::ggsave("vans_service_summaryzw.png", width = 2600, height = 1600, units = "px")
```




## Visualizing Distributions of Service Cost and Sercice times per Engine 

This can be done in a histogram or a violin plot or a boxplot, or a histogram  

```{r visualizing distributions}

# The violin plot is not visually appealing due to the skewness of the data. Therefore, We will use box plots to indicate the skewness of the data

# A histogram for service times distribution

p1 <- ggplot2::ggplot(data = van_data_final, ggplot2::aes(x = mileage)) + 
  ggplot2::geom_histogram(fill = "white", color = "black") +
  ggplot2::geom_density()+
  ggplot2::facet_grid(engine_type~., scales = "free")+
  ggplot2::scale_x_continuous(breaks = seq(0,300000, 30000)) + 
  ggplot2::ylab("Service times count") +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
  ggplot2::labs(title = "Service Times Distribution", subtitle = "A Histogram showing the distribution of service times per engine")

ggplot2::ggsave("service_times_histogram.png", width = 2600, height = 1600, units = "px")

#A box plot for service times distribution
ggplot2::ggplot(data = van_data_final, ggplot2::aes(y = engine_type, x = mileage)) +
  ggplot2::geom_boxplot()+
  ggplot2::theme_bw() +
  ggplot2::scale_y_discrete(limits =rev(c("E150 4.6L", "GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.5L", "TRANSIT 3.7L", "VOYAGER 3.6L")) )+
  ggplot2::scale_x_continuous(breaks = seq(0, 300000, 50000)) + 
 ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
  ggplot2::labs(title = "Service times Distribution", subtitle = "A box plot showing the distribution of service times per engine")


ggplot2::ggsave("service_times_boxplot.png", width = 2600, height = 1600, units = "px")

#A box plot for total service cost distribution

ggplot2::ggplot(data = van_data_final, ggplot2::aes(y = engine_type, x = total_cost)) +
  ggplot2::geom_boxplot()+
  ggplot2::theme_bw() +
  ggplot2::scale_y_discrete(limits =rev(c("E150 4.6L", "GC 3.6L", "PROMASTER 3.6L", "TRANSIT 3.5L", "TRANSIT 3.7L", "VOYAGER 3.6L")) ) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
  ggplot2::labs(title = "Total Cost per Engine Distribution", subtitle = "A box plot showing the distribution of service costs per engine")

ggplot2::ggsave("service_cost_box_plot.png", width = 2600, height = 1600, units = "px")

  
# A histogram for total service cost distribution
ggplot2::ggplot(data = van_data_final, ggplot2::aes(x = total_cost)) + 
  ggplot2::geom_histogram(fill = "white", color = "black") +
  ggplot2::facet_grid(engine_type~., scales = "free")+
  ggplot2::scale_x_continuous(breaks = seq(0,7000, 500)) + 
  ggplot2::ylab("Sercive cost count") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Total Cost per Engine Distribution", subtitle = "A Histogram showing the distribution of service costs per engine")

ggplot2::ggsave("service_cost_histogram.png", width = 2600, height = 1600, units = "px")
```



## Visualizing the Cumulative cost

cumulative costs were calculated per van, per engine

```{r calculating the cumulative  }
# Plotting a curve with cumulative plots
#Using faceting does not offer clarity when it comes to cumulative costs. 
#This is mainly due to the difference in the number of vehicles and consequently due to the number of service times

#Therefore, I will filter the data and represent it per engine

e150_data <- dplyr::filter(van_data_final, engine_type == "E150 4.6L")
gc36_data <- dplyr::filter(van_data_final, engine_type == "GC 3.6L")
promaster_data <- dplyr::filter(van_data_final, engine_type == "PROMASTER 3.6L")
transit35_data <- dplyr::filter(van_data_final, engine_type == "TRANSIT 3.5L")
transit37_data <- dplyr::filter(van_data_final, engine_type == "TRANSIT 3.7L")
voyager_data <- dplyr::filter(van_data_final, engine_type == "VOYAGER 3.6L")



```

### Visualizing cumulative costs for all engines  
This visualization will help show after how many miles replacing the engine becomes too expensive. The average cost of a new engine is between $2000 - $3000. Therefore, the estimated price of a new engine will be set at $2500. 

```{r all vans cumulative cost}

van_line <-  data.frame(
  xend =  c(116050, 153000, 174800, 198400,230590, 116050, 153000, 174800, 198400,230590),
  x = c(116050, 153000, 174800, 198400,230590, 0, 0, 0, 0,0),
  yend = c(2500, 5000, 7500, 10000, 12500, 2500, 5000, 7500, 10000, 12500),
  y = c(0, 0, 0, 0, 0, 2500, 5000, 7500, 10000, 12500)
)
van_annotate <- data.frame(
  xend = c(116050, 25000),
  yend = c(1000, 5000),
  x = c(0, 25000),
  y= c(1000, 2500)
)


ggplot2::ggplot(data = van_data_final, ggplot2::aes(x= mileage, y = overall_cum_cost/32))+
  ggplot2::geom_line()+
  ggplot2::scale_x_continuous(breaks = seq(0, 300000,50000 )) +
  ggplot2::theme_bw()+
  ggplot2::geom_segment(data = van_line, ggplot2::aes(x= x, y= y, xend = xend, yend = yend), linetype = "dashed") +
  ggplot2::geom_segment(data = van_annotate, ggplot2::aes(x= x, y= y, xend = xend, yend = yend)) + #the arrow argument failed to work
  ggplot2::geom_point(data = van_annotate, ggplot2::aes(x= xend, y = yend), shape = 20, size =2 ) +
  ggplot2::geom_point(data = van_annotate, ggplot2::aes(x= x, y = y), shape = 20, size =2 )+
  ggplot2::annotate("text", x= 55000, y = 3500, label = "Price of a new engine", size = 3, ) +
  ggplot2::annotate("text", x = 50000, y = 1500, label = "Miles covered", size = 3) +
  ggplot2::labs(title = "Average Cumulative Cost Vs Miles Covered for All Vans", caption = "The average cumulative cost was calculated by dividing the \n overall cumulative cost per engine by the number of vans per engine")+
  ggplot2::ylab("Average cumulative cost per van($)") + 
  ggplot2::xlab("Miles covered (Miles)")

ggplot2::ggsave("cumulative_cost_all_vans.png", width = 1624, height = 900, units = "px")


```



### Visualizing Cumulative Costs for GC 3.6L engine

This will help estimate the when it becomes too expensive to repair the engine.  
The average cost of a new GC 3.6L engine is $2500.

```{r GC 3.6L cumulative cost}

gc_line <-  data.frame(
  xend =  c(142600, 170671, 196261, 218641,142600, 170671, 196261, 218641),
  x = c(142600, 170671, 196261, 218641, 0, 0, 0, 0),
  yend = c(2500, 5000, 7500, 10000, 2500, 5000, 7500, 10000),
  y = c(0, 0, 0, 0, 2500, 5000, 7500, 10000)
)
gc_annotate <- data.frame(
  xend = c(5000, 142600),
  yend = c(2500, 0),
  x = c(5000, 0),
  y= c(0, 0)
)


ggplot2::ggplot(data = gc36_data, ggplot2::aes(x= mileage, y = cum_cost_per_engine/15))+
  ggplot2::geom_line()+
  ggplot2::scale_x_continuous(breaks = seq(0, 300000,50000 )) +
  ggplot2::theme_bw()+
  ggplot2::geom_segment(data = gc_line, ggplot2::aes(x= x, y= y, xend = xend, yend = yend), linetype = "dashed") + 
  ggplot2::geom_segment(data = gc_annotate, ggplot2::aes(x= x, y= y, xend = xend, yend = yend)) + #the arrow argument failed to work
  ggplot2::geom_point(data = gc_annotate, ggplot2::aes(x= xend, y = yend), shape = 20, size =2 ) +
  ggplot2::geom_point(data = gc_annotate, ggplot2::aes(x= x, y = y), shape = 20, size =2 )+
  ggplot2::annotate("text", x= 30000, y = 1800, label = "Price of a new engine", size = 3, ) +
  ggplot2::annotate("text", x = 100000, y = 500, label = "Miles covered", size = 3) +
  ggplot2::labs(title = "Average Cumulative Cost Vs Miles Covered for GC 3.6L", caption = "The average cumulative cost was calculated by dividing the overall \n cumulative cost per engine by the number of vans per engine")+
  ggplot2::ylab("Average cumulative cost per van($)") + 
  ggplot2::xlab("Miles covered (Miles)")


ggplot2::ggsave("cumulative_cost_gc_36L.png", width = 1624, height = 900, units = "px")

```

### Visualizing Cumulative Costs for GC 3.6L engine

This will help estimate the when it becomes too expensive to repair the engine.  
The average cost of a new promaster 3.6L engine is $3000.

```{r cumulative cost Promaster 3.6L}

promaster_line <-  data.frame(
  xend =  c(102590, 130370, 149800, 158700,174950, 198290,211900, 102590, 130370, 149800, 158700,174950, 198290,211900),
  x = c(102590, 130370, 149800, 158700,174950, 198290,211900, 0, 0, 0, 0, 0, 0, 0),
  yend = c(3000, 6000, 9000, 12000, 15000, 18000, 21000, 3000, 6000, 9000, 12000, 15000, 18000, 21000),
  y = c(0, 0, 0, 0, 0, 0, 0, 3000, 6000, 9000, 12000, 15000, 18000, 21000)
)
promaster_annotate <- data.frame(
  xend = c(102590, 10000),
  yend = c(1000, 6000),
  x = c(0, 10000),
  y= c(1000, 3000)
)


ggplot2::ggplot(data = promaster_data, ggplot2::aes(x= mileage, y = cum_cost_per_engine/8))+
  ggplot2::geom_line()+
  ggplot2::scale_x_continuous(breaks = seq(0, 300000,50000 )) +
  ggplot2::theme_bw()+
  ggplot2::geom_segment(data = promaster_line, ggplot2::aes(x= x, y= y, xend = xend, yend = yend), linetype = "dashed") + 
  ggplot2::geom_segment(data = promaster_annotate, ggplot2::aes(x= x, y= y, xend = xend, yend = yend)) + #the arrow argument failed to work
  ggplot2::geom_point(data = promaster_annotate, ggplot2::aes(x= xend, y = yend), shape = 20, size =2 ) +
  ggplot2::geom_point(data = promaster_annotate, ggplot2::aes(x= x, y = y), shape = 20, size =2 )+
  ggplot2::annotate("text", x= 35000, y = 4500, label = "Price of a new engine", size = 3, ) +
  ggplot2::annotate("text", x = 50000, y = 1500, label = "Miles covered", size = 3) +
  ggplot2::labs(title = "Average Cumulative Cost Vs Miles Covered for Promaster 3.6L", caption = "The average cumulative cost was calculated by dividing the overall \n cumulative cost per engine by the number of vans per engine")+
  ggplot2::ylab("Average cumulative cost per van ($)") + 
  ggplot2::xlab("Miles covered (Miles)")
  

ggplot2::ggsave("cumulative_cost_promaster.png", width = 1624, height = 900, units = "px")
```





### Visualizing Transit 3.7L cumulative cost data  

```{r cumulative cost Transit 3.7L}

transit37_line <-  data.frame(
  xend =  c(85940, 154600, 183000, 209100,269000, 85940, 154600, 183000, 209100,269000),
  x = c(85940,154600, 183000, 209100, 269000, 0, 0, 0, 0, 0),
  yend = c(2500, 5000, 7500, 10000, 12500, 2500, 5000, 7500, 10000, 12500),
  y = c(0, 0, 0, 0, 0, 2500, 5000, 7500, 10000, 12500)
)
transit37_annotate <- data.frame(
  xend = c(25000, 85940), 
  yend = c(5000, 1250),
  x = c(25000, 154600),
  y= c(2500, 1250)
)


ggplot2::ggplot(data = transit37_data, ggplot2::aes(x= mileage, y = cum_cost_per_engine/5))+
  ggplot2::geom_line()+
  ggplot2::scale_x_continuous(breaks = seq(0, 300000,50000 )) +
  ggplot2::theme_bw()+
  ggplot2::geom_segment(data = transit37_line, ggplot2::aes(x= x, y= y, xend = xend, yend = yend), linetype = "dashed") + 
  ggplot2::geom_segment(data = transit37_annotate, ggplot2::aes(x= x, y= y, xend = xend, yend = yend)) + #the arrow argument failed to work
  ggplot2::geom_point(data = transit37_annotate, ggplot2::aes(x= xend, y = yend), shape = 20, size =2 ) + 
  ggplot2::geom_point(data = transit37_annotate, ggplot2::aes(x= x, y = y), shape = 20, size =2 )+
  ggplot2::annotate("text", x= 55000, y = 3500, label = "Price of a new engine", size = 3, ) + 
  ggplot2::annotate("text", x = 125000, y = 1500, label = "Miles covered", size = 3) + 
  ggplot2::labs(title = "Average Cumulative Cost Vs Miles Covered for Transit 3.7L", caption = "The average cumulative cost was calculated by dividing the overall \n cumulative cost per engine by the number of vans per engine")+
  ggplot2::ylab("Average cumulative cost per van") + 
  ggplot2::xlab("Miles covered")


  
ggplot2::ggsave("cumulative_cost_transit_37l.png", width = 1624, height = 900, units = "px")

```

